scRNAseq benchmark in want of a title

### 2019-05-19 bmc_article.tex

%%% Load packages
%\usepackage{amsthm,amsmath}
%\RequirePackage{natbib}
\RequirePackage[authoryear]{natbib}% uncomment this for author-year bibliography
\usepackage{hyperref}
\usepackage[utf8]{inputenc} %unicode support
%\usepackage[applemac]{inputenc} %applemac support if unicode package fails
%\usepackage[latin1]{inputenc} %UNIX support if unicode package fails
\usepackage{graphicx}

@@

\subsubsection*{Excluded more cells is not necessarily good}

@@

A first observation was that very stringent filtering (removing up to 20\% of the cells) was not generally associated with an increase in accuracy of the clustering (Supplementary Figure 8A). Moreover, stringent filtering tended to be biased towards the exclusion of specific subpopulations (Supplementary Figure 8B). Indeed, applying the same filtering criteria on individual clusters of cells (identified through scran's `quickCluster` method) resulted in nearly no cell being filtered out, which suggests that filtering on the global population tends to discard cells of subpopulations with more extreme properties, rather than `bad` cells.

@@

We next investigated the impact of normalization on the ability to distinguish subpopulations, which was much greater than that of filtering (Supplementary Figure 13). For many clusters, scaling tended to reduce both the average silhouette width (Supplementary Figure 12B) and the proportion of variance in the top principal components explained by subpopulations (Figure \ref{fig:figure3} and Supplementary Figure 13), and was associated with a poorer adjusted Rand index (ARI) of the resulting clustering (Figure \ref{fig:figure3}). Regressing out covariates systematically gave poorer performance on all metrics. sctransform produced the highest cluster average silhouette width (Supplementary Figure 12B) as well as ARI (Figure \ref{fig:figure3}), but not always the highest mean proportion of variance in the top components explained by subpopulations (Figure \ref{fig:figure3} and Supplementary Figure 13). Interestingly, clusters with the lowest average silhouette width (and which are therefore more difficult to classify) tended to be less well distinguished with sctransform, and instead showed a good difference in non-normalized and/or non-scaled data, suggesting that some procedures emphasize differences more useful to track some subpopulations than others, and hence the possibility that further developments in methods could improve this step.

@@

All analyses were performed through the \url[pipComp]{https://github.com/plger/pipComp} R package, which implements the pipeline framework described here, and all code to reproduce each figure is available in the \url[https://github.com/markrobinsonuzh/scRNA\_pipelines\_paper]{https://github.com/markrobinsonuzh/scRNA_pipelines_paper} repository, which also includes the basic datasets with a standardized annotation.

\subsection*{How the data was obtained}
(to come)

\subsection*{Software and package versions}
(to come)

@@

\subsection*{Doublet detection method}
Our new doublet detection method is available at \url{https://github.com/plger/plDoubletFinder}. We first cluster the cells using scran's \textit{quickCluster} method, favouring overclustering, and reduce the dataset to the union of 1500 most expressed genes of each cluster. We then create 75\% of the artifical doublets by sampling the two cells specifically from different clusters, and 25\% by entirely random sampling in case the clustering was erroneous, and summing their counts. We also create meta-cells from each cluster, and use them to create additional doublets and triplets. We then combine them with the real cells, convert the counts to scaled column ranks, and build a graph of the \textit{k} (default 10) nearest neighbors using scran. We then calculate, for each cell, the proportion of its neighbors that are artificial doublets. This ratio serves as a doublet score, which is then thresholded by simultaneously minimizing the error in classifying real vs artificial doublets and the deviation from the distribution of expected doublet rate (accounting for homotypic doublets as done by DoubletFinder). For the pipeline comparison, the expected doublet rate was set to 2.1\% (with a 1\% standard deviation) on the basis of the mixology datasets.

@@

Unless specified otherwise, the variance in gene expression explained by subpopulations was calculated on the data normalized and transfored through sctransform. For each gene, we fitted a linear model using the subpopulation as only independent variable (\textasciitilde subpopulation), and used the R-squared as a measure of the variance explained. The same method was used for principal components.

@@

@@

\begin{table}[h!]
\caption{Description of the benchmark datasets (GSEids to be added)}
\label{tab:table1}
\begin{tabular}{rlrrl}
  \hline
dataset & protocol & nbCells & subpops & description \\ 
  \hline
Koh & SMARTer & 531 &   9 & FACS purified H7 hESC in different differention stages \\ 
  Kumar & SMARTer & 246 &   3 & Mouse ESC cultured in different conditions \\ 
  Zhengmix4eq & 10x & 3994 &   4 & Mixtures of FACS purified PBMCs \\ 
  Zhengmix4uneq & 10x & 6498 &   4 & Mixtures of FACS purified PBMCs \\ 
  Zhengmix8eq & 10x & 3994 &   8 & Mixtures of FACS purified PBMCs \\ 
  mixology10x3cl & 10x & 902 &   3 & Mixture of 3  cancer cell lines from CellBench \\ 
  mixology10x5cl & 10x & 3918 &   5 & Mixture of 5 cancer cell lines from CellBench \\ 
   \hline
\end{tabular}
\end{table}

>>>>

%%% Load packages
%\usepackage{amsthm,amsmath}
%\RequirePackage{natbib}
\RequirePackage[authoryear]{natbib}% uncomment this for author-year bibliography
\usepackage{hyperref}
\usepackage[utf8]{inputenc} %unicode support
%\usepackage[applemac]{inputenc} %applemac support if unicode package fails
%\usepackage[latin1]{inputenc} %UNIX support if unicode package fails
\usepackage{graphicx}
\usepackage{url}

@@

\subsubsection*{Excluding more cells is not necessarily good}

@@

A first observation was that very stringent filtering (removing up to 20\% of the cells) was not generally associated with an increase in accuracy of the clustering (Supplementary Figure 8A). Moreover, stringent filtering tended to be biased towards the exclusion of specific subpopulations (Supplementary Figure 8B). Indeed, applying the same filtering criteria on individual clusters of cells (identified through scran's \textit{quickCluster} method) resulted in nearly no cell being filtered out, which suggests that filtering on the global population tends to discard cells of subpopulations with more extreme properties, rather than `bad` cells.

@@

We next investigated the impact of normalization on the ability to distinguish subpopulations, which was much greater than that of filtering (Supplementary Figure 13). For many clusters, scaling tended to reduce both the average silhouette width (Supplementary Figure 12B) and the proportion of variance in the top principal components explained by subpopulations (Figure \ref{fig:figure3} and Supplementary Figure 13), and was associated with a poorer adjusted Rand index (ARI) of the resulting clustering (Figure \ref{fig:figure3}). Regressing out covariates systematically gave poorer performance on all metrics. sctransform produced the highest cluster average silhouette width (Supplementary Figure 12B) as well as ARI (Figure \ref{fig:figure3}), but not always the highest mean proportion of variance in the top components explained by subpopulations (Figure \ref{fig:figure3} and Supplementary Figure 13). Of note, even though sctransform is meant to be applied to data with unique molecular identifiers (UMI), it also performed fairly well with smartseq datasets. Interestingly, clusters with the lowest average silhouette width (and which are therefore more difficult to classify) tended to be less well distinguished with sctransform, and instead showed a good difference in non-normalized and/or non-scaled data, suggesting that some procedures emphasize differences more useful to track some subpopulations than others, and hence the possibility that further developments in methods could improve this step.

@@

\subsection*{Code and data availability}

All analyses were performed through the \href{pipComp}{https://github.com/plger/pipComp} R package, which implements the pipeline framework described here. All code to reproduce each figure is available in the \url{https://github.com/markrobinsonuzh/scRNA\_pipelines\_paper} repository, which also includes the basic datasets with a standardized annotation.

The gene counts for the two mixology datasets were downloaded from the  \href[CellBench repository]{https://github.com/LuyiTian/CellBench\_data/data/sincell\_with\_class.RData}, commit 74fe79e. For the other datasets, we used the unfiltered counts from \citep{duoClustering2018}, available on the corresponding repository \url{https://github.com/markrobinsonuzh/scRNAseq\_clustering\_comparison}. For simplicity, all starting \textit{SingleCellExperiment} objects with standardized metadata are available on \url{https://github.com/markrobinsonuzh/scRNA\_pipelines\_paper}.

\subsection*{Software and package versions}
Analyses were performed in R 3.6.0 (bioconductor 3.9), and the following packages were installed from github repositories: Seurat (version 3.0.0), sctransform (0.2.0), DoubletFinder (2.0.1), scds (1.0.0), pipComp (0.2). The code for the glmPCA was obtained from \url{https://github.com/willtownes/scrna2019} (commit 1ddcc30ebb95d083a685f12fe81d35dd1b0cb1b2).

@@

\subsection*{Doublet detection method}
Our doublet detection method is available at \url{https://github.com/plger/plDoubletFinder}. Briefly, we first cluster the cells using scran's \textit{quickCluster} method, favouring overclustering (default minimum size of 20), and reduce the dataset to the union of 1500 most expressed genes of each cluster. We then create 75\% of the artifical doublets by sampling the two cells specifically from different clusters, and 25\% by entirely random sampling (in case the clustering was erroneous), and summing the counts of each pair of cells. We also create meta-cells from each cluster, and use them to create additional doublets and triplets. We then combine them with the real cells, convert the counts to scaled column ranks, and build a graph of the \textit{k} (default 10) nearest neighbors using scran. We then calculate, for each cell, the proportion of its neighbors that are artificial doublets. This ratio serves as a doublet score, which is then thresholded by simultaneously minimizing the error in classifying real vs artificial doublets and the deviation from the distribution of expected doublet rate (accounting for homotypic doublets as done by DoubletFinder). For the pipeline comparison here, the expected doublet rate was set to 2.1\% (with a 1\% standard deviation) on the basis of the mixology datasets.

@@

Unless specified otherwise, the variance in gene expression explained by subpopulations was calculated on the data normalized and transformed through sctransform. For each gene, we fitted a linear model using the subpopulation as only independent variable (\textasciitilde subpopulation), and used the R-squared as a measure of the variance explained. The same method was used for principal components.

@@

To estimate the correlation between the first principal components and covariates such as library size, we first fitted a linear model on the subpopulations, and correlated the residuals of this model with the covariate of interest.

@@

\begin{table}[h!]
\caption{Overview of the benchmark datasets}
\label{tab:table1}
\begin{tabular}{rlrrl}
  \hline
dataset & source & protocol & nbCells & subpops & description \\ 
  \hline
Koh & GSE85066 & SMARTer & 531 &   9 & FACS purified H7 hESC in different differention stages \\ 
  Kumar & GSE60749 & SMARTer & 246 &   3 & Mouse ESC cultured in different conditions \\ 
  Zhengmix4eq & \citep{duoClustering2018} & 10x & 3994 &   4 & Mixtures of FACS purified PBMCs \\ 
  Zhengmix4uneq & \citep{duoClustering2018} & 10x & 6498 &   4 & Mixtures of FACS purified PBMCs \\ 
  Zhengmix8eq & \citep{duoClustering2018} & 10x & 3994 &   8 & Mixtures of FACS purified PBMCs \\ 
  mixology10x3cl & \cite{tianMixology2018} & 10x & 902 &   3 & Mixture of 3  cancer cell lines from CellBench \\ 
  mixology10x5cl & \cite{tianMixology2018} & 10x & 3918 &   5 & Mixture of 5 cancer cell lines from CellBench \\ 
   \hline
\end{tabular}
\end{table}
